\bf Adaption of the Chumbley Score to matching of bullet striation marks

### 2018-04-16 doc/paper.Rmd
data uploaded

Comparisons of striae from bullets are usually based on comparisons of striae in land engraved areas, which are extracted in form of cross sections, called *profiles* \citep{aoas,ma2004}. From profiles bullet *signatures* \citep{chu2013,aoas} are extracted as residuals of a loess fit or Gaussian filter. This effectively removes topographic structure from the data in the attempt to increase the signal to noise ratio. The span of the loess fit was found using cross-validation, as described by \citet{aoas}.
 <!-- Signatures are considered to be noise free and a good reflection of the key attributes of the raw marking, and the unique features of a bullet. \hh{Do you have a reference for the previous sentenced?} A  detailed discussion of the extraction technique for signatures is given in \citet{aoas}. -->
 <!--, where comprehensive details about the height at which profile is to be selected, removing curvature, smoothing, identifying groove locations are explained.-->

@@

<!--
## Potential Challenges in Chumbley Score Adaptation

Markings made by screw driver tips \citep{chumbley} are longer and pronounced and about 7 mm in length \citep{manytoolmarks1}, while the groove to groove length of the bullet lands from the Hamby study \citep{hamby} are about 2mm in length. The Ruger pistol barrels used in the  Hamby study have a 9mm caliber. This means that bullets used in these pistols have land markings of only about a quarter of the size of screwdriver toolmarks used by \citet{chumbley, hadler}, therefore raising the problem of potential loss in power of the test. -->

<!--One way to investigate this is to evaluate bullet striations at different resolutions. This gives us an incentive to use both the NIST and Iowa State scans and investigate the effect of length of the digitized mark on the working of the algorithm. Apart from this, the algorithm developed by \citet{hadler} has only been tested for flatter and wider surfaces which have negligible curvature, whereas striations on bullets are made on their curved surfaces. Therefore, using methods proposed for toolmarks need adaptation in order to give tangible results for bullets. Another challenge that arises from these topographical differences, is the understanding of that effect of different levels of smoothing. Especially investigating its effect on the algorithm performance when signatures are being compared and when raw markings or profiles are being compared.

Some other challenges associated with the adaptation process call for a brief understanding of the methodology involved. The Chumbley method and algorithm works on the premise of comparison windows that are supposed to represent small segments of pixels of equal sizes in the two markings.  These windows are to be compared in a predetermined fashion and are termed as windows of optimization and validation. The optimization window is selected in the first step called the optimization step where areas/windows of best agreement between the the two markings are found.

The window of optimization for bullet striations are shorter as bullet signatures in terms of absolute length (in mm), are smaller compared to toolmarks. Even in terms of digitized pixels, there is no straightforward way of establishing what window size is to be chosen. One important way to address this is to keep the number of windows of optimization sufficiently large or not too small. This means, if the number of windows are too small we inadvertently use bigger window sizes as we want to compare all segements of one mark with the segments of the other mark. This leads to not giving enough emphasis to smaller segements of the two marking that might be in high agreement. In terms of absolute length we end up comparing smaller segments between the two markings while dealing with bullet lands, than those being compared for toolmarks.

For too large window sizes, the weight of small features that would otherwise uniquely classify a signature and hence identify the region of agreement is vastly reduced. On the other hand, this introduces another problem as, if we go too small in the window of optimization, the unique features of the trace segments are lost and all seem similar. The problem of distinctiveness in the structure of bullet striations may also lead to potential failure of the algorithm to identify correct windows of agreements or not be able to identify these windows at all. This may lead to incorrect classification of bullets as coming from the same-source (or matches) or from different source (or non-matches).

Thus, the comaparison windows have a direct influence on the false positives and false negatives, making identification of the optimum sizes of these windows very important. This raises important questions about how and what parameter settings need to be chosen for bullet land comparison. Something similar is done in \citet{afte-chumbley} for toolmark comparisons of slip-joint pliers where optimum window sizes are determined. There is a need to figure out the best parameter settings which minimizes the errors for unsmoothed markings or profiles and pre-processed signatures. An analysis of these error rates and comparison with other methods will help us understand the adaptibility of the chumbley score to bullets.  
-->

@@

## Setup

@@

@@

wo120_1<- wo120 %>% gather(type, error, beta) %>%  
  ggplot(aes(x = wv, y = error, colour = factor(alpha)))+
  geom_line(linetype=5)+  geom_point(shape= 24) + facet_wrap(~type, scales="free") +
  theme_bw()+ 
  theme(legend.position="bottom") + ggtitle("Signatures" , subtitle = "Varying validation window sizes, Optimization window = 120")

combined120<- 
  ggplot(data = wo120 %>% gather(type, error, beta), aes(x = wv, y = error, colour = factor(alpha)))+
  geom_line(aes(linetype="Signatures"))+  geom_point(shape= 24) + 
  geom_line(data =  errors2 %>% filter(wo==120, coarse == 0.25) %>% 
  gather(type, error, beta), aes(x = wv, y = error, colour = factor(alpha), linetype = "Profiles")) +       geom_point(data =  errors2 %>% filter(wo==120, coarse == 0.25) %>% 
  gather(type, error, beta)) +
  facet_wrap(~type, scales="free") +
  scale_linetype_manual("Lines",values=c("Signatures"=2,"Profiles"=1))+
  guides(fill = guide_legend(keywidth = 1, keyheight = 1),
    linetype=guide_legend(keywidth = 3, keyheight = 1),
    colour=guide_legend(keywidth = 3, keyheight = 1))+
  theme_bw()+ 
  theme(legend.position="right") + ggtitle("Type II error with Validation window" , subtitle = "Profiles at coarseness 0.25") + xlab("Type II error")

@@

<!-- ### other questions -->
<!-- How much movement is reasonable and should be used? Is there a way to understand if the 2nd marking in comparison to the 1st marking is compressed or elongated or neither? Should the amount of movement selected depend on this compression or elongation? or using a static delta movement value justifieable. -->

# Appendix
\begin{appendix}

@@

```{r}
#errors %>% ggplot(aes(x = wv, y=total, colour=factor(alpha), shape="signatures")) + geom_point() + facet_grid(~wo) +geom_point(data=errors2, aes(shape="profiles"))

@@

total2<- wo120 %>% filter(wv>10) %>% ggplot(aes(x = wv, y=total, colour=factor(alpha), shape="signatures")) + geom_point() + geom_line() +geom_point(data=errors2 %>% filter(wv>10) %>% filter(coarse==0.25), aes(shape="profiles")) + geom_line(data=errors2 %>% filter(wv>10)%>% filter(coarse==0.25), aes(shape="profiles")) + theme_bw() +
  scale_colour_brewer(expression("Nominal\ntype I error"~alpha), palette="Set2") + ylab("Total error") + ggtitle(label = "(e)")

@@

#multiplot(plotlist = list(total1, total2, miss_1, miss_2, e1, e2), cols = 2)#, matrix(c(1,2,3,3), nrow=2, byrow=TRUE))
multiplot(plotlist = list(e1, total1, total2,e2, miss_1, miss_2), cols = 2)
```
\end{appendix}

>>>>

Comparisons of striae from bullets are usually based on comparisons of striae in land engraved areas, which are extracted in form of cross sections, called *profiles* \citep{aoas,ma2004}. From profiles bullet *signatures* \citep{chu2013,aoas} are extracted as residuals of a loess fit or Gaussian filter. This effectively removes topographic structure from the data in the attempt to increase the signal to noise ratio. The span of the loess fit was found using cross-validation, as described by \citet{aoas}.

@@

@@

## Setup
\hh{XXX the description below needs to be updated.}

@@

# Results

\hh{Figure \ref{fig:coarse} shows the type II error rates for profiles using an optimization window $w_o= 120$ and a validation window $w_v = 30$ for varying level of coarseness. The type II error for all nominal levels of  $\alpha$ are the lowest for a coarseness range of 0.20 to 0.30, with a minim in type II reached at a coarseness $c = 0.25$. }
\hh{However, for an optimal value of coarseness other aspects have to be condsidered.}
```{r coarse, fig.cap="Type II error with respect to coarseness parameter over profiles, $w_o = 120$, $w_v = 30$. Optimal values for coarseness are around $c \\approx 0.30$", fig.width=5, fig.height = 3, out.width='.6\\textwidth'}
e1<-errors2 %>% filter(wv==30)%>% ggplot(aes(x = coarse, y=beta, colour=factor(alpha))) +
  geom_point() + geom_line() +
  theme_bw() +
  scale_colour_brewer(expression("Nominal\ntype I error"~alpha), palette="Set2") + xlab("Profile Coarseness")+
  ylab("Type II error") 
e1
```

profc20 <- readRDS("../data/profiles-coarseness/chumbley-csafe-profiles-inv-wo:160-wv:30-c:20.rds")
profc20$coarseness <- 0.2
profc25 <- readRDS("../data/profiles-coarseness/chumbley-csafe-profiles-inv-wo:160-wv:30-c:25.rds")
profc25 <- readRDS("../data/profiles-coarseness/chumbley-csafe-profiles-inv-wo:160-wv:50-c:25.rds")
profc30 <- readRDS("../data/profiles-coarseness/chumbley-csafe-profiles-inv-wo:160-wv:30.rds")
profc30$coarseness <- .30
profc <- bind_rows(profc20, profc25, profc30)
profc <- profc %>% mutate(
  locations = chumbley %>% purrr::map_int(.f = function(x) {
    res <- NULL
    try(res <- x$locations[1], silent = TRUE)
    res
  })
)
profsumm <- read.csv("../data/profiles-summary.csv")
profsumm <- profsumm %>% select(-run_id)
profc <- profc %>% left_join(profsumm, by=c("land2_id"="land_id", "profile2_id"="profile_id"))

@@

wo120_1<- wo120_profile %>% gather(type, error, beta) %>%  
  ggplot(aes(x = wv, y = error, colour = factor(alpha)))+
  geom_line(linetype=5)+  geom_point(shape= 24) + facet_wrap(~type, scales="free") +
  theme_bw()+ 
  theme(legend.position="bottom") + ggtitle("Signatures" , subtitle = "Varying validation window sizes, Optimization window = 120")

combined120<- 
  ggplot(data = wo120_profile %>% gather(type, error, beta), aes(x = wv, y = error, colour = factor(alpha)))+
  geom_line(aes(linetype="Signatures"))+  geom_point(shape= 24) + 
  geom_line(data =  errors2 %>% filter(wo==120, coarse == 0.25) %>% 
  gather(type, error, beta), aes(x = wv, y = error, colour = factor(alpha), linetype = "Profiles")) +       geom_point(data =  errors2 %>% filter(wo==120, coarse == 0.25) %>% 
  gather(type, error, beta)) +
  facet_wrap(~type, scales="free") +
  scale_linetype_manual("Lines",values=c("Signatures"=2,"Profiles"=1))+
  guides(fill = guide_legend(keywidth = 1, keyheight = 1),
    linetype=guide_legend(keywidth = 3, keyheight = 1),
    colour=guide_legend(keywidth = 3, keyheight = 1))+
  theme_bw()+ 
  theme(legend.position="right") + ggtitle("Type II error with Validation window" , subtitle = "Profiles at coarseness 0.25") + xlab("Type II error")

@@

<!-- ### other questions -->
<!-- How much movement is reasonable and should be used? Is there a way to understand if the 2nd marking in comparison to the 1st marking is compressed or elongated or neither? Should the amount of movement selected depend on this compression or elongation? or using a static delta movement value justifieable. -->
<!--
# Appendix
\begin{appendix}

@@

```{r, eval=FALSE}
#errors %>% ggplot(aes(x = wv, y=total, colour=factor(alpha), shape="signatures")) + geom_point() + facet_grid(~wo) +geom_point(data=errors2, aes(shape="profiles"))

@@

total2<- wo120_profile %>% filter(wv>10) %>% ggplot(aes(x = wv, y=total, colour=factor(alpha), shape="signatures")) + geom_point() + geom_line() +geom_point(data=errors2 %>% filter(wv>10) %>% filter(coarse==0.25), aes(shape="profiles")) + geom_line(data=errors2 %>% filter(wv>10)%>% filter(coarse==0.25), aes(shape="profiles")) + theme_bw() +
  scale_colour_brewer(expression("Nominal\ntype I error"~alpha), palette="Set2") + ylab("Total error") + ggtitle(label = "(e)")

@@

#multiplot(plotlist = list(total1, total2, miss_1, miss_2, e1, e2), cols = 2)#, matrix(c(1,2,3,3), nrow=2, byrow=TRUE))
multiplot(plotlist = list(e1, total1, total2,e2, miss_1, miss_2), cols = 2)
```
\end{appendix} 
-->

### 2018-04-16 doc/paper.tex
data uploaded

@@

\subsection{Results}\label{results}

@@

\section{Results}\label{results-1}

\hh{XXX We have been operating under a false assumption: smoothing the signature at f=1 DOES change it - see figure \ref{fig:profile-sketch}. We need to re-run some of the results }

\caption{Overview of the effect of different coarseness parameters $c$ on the profile shown in Figure \ref{fig:sigs-profiles} (top). The bottom row shows histograms of  the optimal locations $t^o$ identified in the optimization step for different values of the coarseness parameter $c$. }\label{fig:profile-sketch}
\end{figure}

\hh{Figure \ref{fig:coarse} shows the type II error rates for profiles using an optimization window $w_o= 120$ and a validation window $w_v = 30$ for varying level of coarseness. The type II error for all nominal levels of  $\alpha$ are the lowest for a coarseness range of 0.20 to 0.30. For the remainder of the analysis we use a fixed value of coarseness of 0.25. }

@@

\section{Appendix}\label{appendix}

\begin{appendix}

On the other hand Figure \ref{fig:prof_missings} (b) shows if the coarseness level set in the chumbley agorithm has any effect on the signatures, which are pre-processed and already smoothed to a certain extent. From Figure \ref{fig:prof_missings} (b) we can notice that for different nominal $\alpha$ levels, the type II error fluctuates slightly but does not change much, thereby helping us conclude that the coarseness levels set in the LOWESS smoothing in the chumbley alggorithm does effect the type II error much for signatures.

\subsection{Comparison of profiles and signatures}

Another reason for failed tests can be incorrect identification of maximum correlation windows in the optimization step as seen in figure \ref{fig:prof_missings}(d) because of the level of smoothing, as too much smoothing would subdue intricate features that might otherwise help in the correlation calculations and correct identification of maximum correlation windows irrespective of the size. This would again cause a simiar effect as explained for figure \ref{fig:missings} with validation windows, irrespective of size, during the shifts end up at the ends of the markings resulting in an invalid calculation and failed comparison attempt.

In figure \ref{fig:prof_missings}(d) and (f), we compare profiles and signatures on the basis of number of failed tests. The profiles chosen for figure \ref{fig:prof_missings}(f) have a constant coarseness of 0.25 and window of optimization as 120. The signatures in this case are not smoothed using the chumbley algorithm step of LOWESS smoothing. Instead signatures are used as calculated by \citet{aoas}. The smoothing in these signatures were determined and fixed on the basis of their performance in the random forest based algorithm proposed by \citet{aoas}. The comparison of profiles and signatures with variation of validation window size therefore is made on even footing. The trends are similar to figure \ref{fig:missings} in the sense that for known non-matches the number of failed tests are more for both signatures and profiles and increasing linearly with the validation window size. The problem is however, worse for profiles which has higher number of failed tests than signatures for all validation windows.

The total error for different validation window sizes for signatures and profiles can be seen in figure \ref{fig:prof_missings} (e).The optimization window size is 120 and profiles are calculated at a default 0.25 coarseness level while signatures as before are not smoothed again in the modified chumbley algorithm. We can see that the total error is always higher for profiles as compared to signatures for all sizes of validation window.

{\centering \includegraphics[width=\textwidth]{figures/prof_missings-1}

\caption{Row 3:  Total error and Number of failed tests by the window validation size, wv, and ground truth, Row 2: Total error and Number of failed tests with Coarseness for both profiles and signatures, Row 1: Type II error for different coarseness levels as used in the modified chumbley algorithm for profiles and signatures}\label{fig:prof_missings}
\end{figure}
\end{appendix}

>>>>

\hh{XXX the description below needs to be updated.}

@@

\section{Results}\label{results}

\hh{Figure \ref{fig:coarse} shows the type II error rates for profiles using an optimization window $w_o= 120$ and a validation window $w_v = 30$ for varying level of coarseness. The type II error for all nominal levels of  $\alpha$ are the lowest for a coarseness range of 0.20 to 0.30, with a minim in type II reached at a coarseness $c = 0.25$. }
\hh{However, for an optimal value of coarseness other aspects have to be condsidered.}

\hh{\autoref{fig:profile-sketch} gives an overview of the effect of different coarseness parameters: from left to right, coarseness levels $c$ are 0.2, 0.25, and 0.3, respectively. The top row shows the resulting signature after smoothing the profile shown in Figure~\ref{fig:sigs-profiles} with the coarseness specified. The histograms in the bottom row show the relative optimal location $t^o$ (optimal location $t^o$ divided by the length of the corresponding profile). Optimal locations should be distributed uniformly. However, for coarseness values of $c = 0.25$ and $c=0.30$ we see very distinct boundary effects: optimal locations $t^o$ are found at the very extreme ends of a profile more often than one would expect based on a uniform location.}

\caption{Overview of the effect of different coarseness parameters $c$ on the profile shown in Figure \ref{fig:sigs-profiles} (top). The bottom row shows histograms of  the (relative) optimal locations $t^o$ identified in the optimization step for different values of the coarseness parameter $c$. }\label{fig:profile-sketch}
\end{figure}

\subsection{Results}\label{results-1}

@@

@@
